---
title: "STA490 EDA"
format: pdf
editor: visual
---

### Data Source

This study will be using the data from the University of Texas Health Science Center at Houston, which is used for research on how MicroRNAs (miRNAs) -138 suppresses glioblastoma proliferation through downregulation. Specifically, miRNAs are small noncoding RNAs implicated in the development, progression, and metastasis of various human cancers. The expression levels of specific miRNAs are used to identify their roles as either tumor suppressors or oncogenes in cancer. The researchers have already cleaned the data, and the data given to us is already cleaned. However, we are uncertain about the data collection process or the data cleaning that they conducted.

### Data Dimensions

The cleaned dataset consists of 627 different miRNAs (rows) and data from 4 healthy individuals and 9 glioblastoma patients (columns). The data has two types of variables: numerical and categorical. The name of each miRNA is coded as text, which is considered a categorical variable. The values that represent expression levels for the control group and the glioblastoma groups are all integers.

### Validation of the Data Cleaning

We have checked that the data does not contain any duplicate miRNA; in other words, each row represents a unique miRNA. Moreover, we also verified that the values for the control and glioblastoma columns are all non-negative, and there are no missing values in this dataset. In this EDA section, there is no distinct outlier since the expression levels of a miRNA vary a lot. Hence, no data was removed, and we did not transform the data.

### Descriptive Statistics

```{r, warning=FALSE, echo=FALSE, message=FALSE}
# Load necessary libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(stringr)
library(knitr)
library(kableExtra)
library(igraph)
library(ggraph)
library(tidyverse)
library(pheatmap)
library(tidyr)
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("DESeq2")
library("DESeq2")
# Load the data
data <- 
  read_excel("/Users/yuqinghu/Desktop/STA490/CollaborateProject/GSE156937.xlsx", 
             sheet = "in")

# Convert miRNA_ID to a factor
data$miRNA_ID <- as.factor(data$miRNA_ID)
```

```{r, warning=FALSE}
# 1. Summary Statistics for Numerical Variables
summary_stats <- data %>% 
  select(-miRNA_ID) %>% 
  summarise(
    mean = colMeans(.),
    median = apply(., 2, median),
    sd = apply(., 2, sd),
    min = apply(., 2, min),
    max = apply(., 2, max)
  ) %>% 
  t() %>% 
  as.data.frame()

# Add row names as a column
summary_stats <- cbind(Statistic = rownames(summary_stats), summary_stats)
rownames(summary_stats) <- NULL

# Rename columns for clarity
colnames(summary_stats) <- c("Statistic", colnames(data)[-1])

# Split the data into two tables
control_table <- summary_stats %>% select(Statistic, Ctrl1, Ctrl2, Ctrl3, Ctrl4, GBM1, GBM2, GBM3)
gbm_table <- summary_stats %>% select(Statistic, GBM4, GBM5, GBM6, GBM7, GBM8, GBM9)

# Generate the first table with the caption
control_table %>%
  kbl(caption = "Summary Statistics for Numerical Variables") %>%
  kable_styling(full_width = FALSE, latex_options = "scale_down")

# Generate the second table without a caption
gbm_table %>%
  kbl() %>%
  kable_styling(full_width = FALSE, latex_options = "scale_down")
```

As the descriptive statistics shown, significant variability in miRNA expression levels between control and GBM samples is revealed with GBM exhibiting greater heterogeneity, as indicated by higher standard deviations (e.g., GBM4 with \~19,286). The median expression values are much lower than the means across all samples, suggesting a right-skewed distribution due to a few highly expressed miRNAs. The maximum expression levels in GBM samples (e.g., GBM4 at \~474,457) underscore the extreme upregulation of certain miRNAs, consistent with the known heterogeneity and adaptability of glioblastoma cells. These findings support the potential critical role of miRNAs in regulating glioblastoma stem cells (GSCs), influencing tumor growth, survival, and therapy resistance. This variability highlights miRNAs as potential therapeutic targets, warranting further exploration to identify specific miRNAs that could disrupt GSC functions and improve GBM treatment outcomes.

```{r, warning=FALSE}
# 2. Distribution of Control vs GBM Values
control_columns <- grep("Ctrl", colnames(data), value = TRUE)
gbm_columns <- grep("GBM", colnames(data), value = TRUE)

data_distribution <- data %>% 
  select(all_of(control_columns), all_of(gbm_columns)) %>% 
  reshape2::melt() %>% 
  mutate(Group = ifelse(grepl("Ctrl", variable), "Control", "GBM"))

# Plot the distribution without outliers and limit y-axis
ggplot(data_distribution, aes(x = Group, y = value, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +
  theme_minimal() +
  labs(title = "Expression Distribution: Control vs GBM", 
       x = "Group", y = "Expression") +
  coord_cartesian(ylim = c(0, 350)) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, 
               color = "red", fill = "red")


```

In this analysis, the data was divided into two groups: **Control** and **GBM**, by selecting the columns labeled "Ctrl" and "GBM" respectively. The expression values were reshaped into a long format to facilitate comparison. A boxplot was generated to visualize the distribution of expression levels for each group, excluding outliers for a clearer representation. The y-axis range was restricted to 0–350 to focus on the primary distribution and exclude extreme values. Red points were added to represent the mean expression for each group.

The boxplot reveals that **Control samples** generally exhibit higher variability in expression levels compared to GBM samples, as evidenced by the wider interquartile range (IQR) for the Control group. The median expression levels appear similar between the two groups, but the GBM group shows a slightly tighter distribution. These findings suggest that GBM cells have less variation in miRNA expression within the observed range, potentially reflecting their more homogeneous cellular behavior compared to the heterogeneous Control group.

```{r}
# 3. Correlation Analysis between Numerical Variables
correlation_matrix <- data %>% 
  select(-miRNA_ID) %>% 
  cor()

# Visualize the Correlation Matrix
heatmap(as.matrix(correlation_matrix), 
        main = "Control & GBM Correlation Matrix", 
        col = topo.colors(10))
```

The correlation matrix reveals distinct clustering of control and GBM samples, with each group showing strong intra-group correlations, as indicated by the prevalence of warmer colors (yellow) within their respective clusters. The weaker inter-group correlations, represented by cooler colors (green/blue), highlight the distinct miRNA expression profiles between control and GBM samples. Within the GBM group, there is evidence of heterogeneity, as some samples show slightly varied correlation patterns, reflecting the biological diversity of glioblastoma cells. These findings underscore the unique regulatory roles of miRNAs in GBM and their potential utility in distinguishing GBM-specific expression profiles from controls.

```{r, warning=FALSE}
# Handle row names
if (!is.null(rownames(data)) && all(!is.na(rownames(data)))) {
  # Clean row names by removing any suffix like "|0"
  rownames(data) <- str_remove(rownames(data), "\\|.*")
} else {
  data <- data %>% column_to_rownames(var = "miRNA_ID")
}

# Validate row names: Remove rows with missing or duplicate row names
data <- data %>%
  filter(!is.na(rownames(data)))  # Ensure no missing row names
rownames(data) <- make.unique(rownames(data))  # Ensure uniqueness

# Ensure all data is numeric
data <- data %>% mutate(across(everything(), as.numeric))

# Validate data: Remove rows with missing values
data <- na.omit(data)

# Transpose the data for correlation across miRNAs
data_t <- as.data.frame(t(data))
```

```{r}

# Compute correlation matrix for miRNAs (rows)
miRNA_correlation_matrix <- cor(data_t, use = "pairwise.complete.obs")

# Convert correlation matrix to a tidy format
correlation_data <- as.data.frame(as.table(miRNA_correlation_matrix)) %>% 
  filter(Var1 != Var2)  # Exclude self-correlations

strongest_correlations <- correlation_data %>% 
  mutate(abs_value = abs(Freq)) %>% 
  arrange(desc(abs_value)) %>% 
  slice(1:25) %>% 
  arrange(desc(Freq))


# Extract unique variables involved in top correlations
unique_vars <- unique(c(strongest_correlations$Var1, 
                        strongest_correlations$Var2))

# Prepare the subset correlation matrix
heatmap_matrix <- miRNA_correlation_matrix[unique_vars, unique_vars]

# Create the heatmap
pheatmap(
  heatmap_matrix,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = FALSE,
  main = "Top 25 Strongest miRNA Correlations",
  color = colorRampPalette(c("white", "orange", "brown"))(50)
)
```

The heatmap displays the top 25 strongest correlations between miRNAs, with correlation coefficients ranging from approximately 0.9992 to 1. Hierarchical clustering organizes miRNAs into groups based on their correlation patterns, indicating potential shared regulatory mechanisms or biological functions. The dominance of warm tones across the heatmap suggests consistently high co-expression among these miRNAs, highlighting their tightly interconnected nature. This strong correlation may imply that these miRNAs are co-regulated or participate in similar pathways, making them promising candidates for further exploration in glioblastoma biology or therapeutic targeting.

```{r}
# Summary table for correlations
correlation_summary <- correlation_data %>% 
  summarise(
    Positive = sum(Freq > 0),
    Negative = sum(Freq < 0),
    Mean = mean(Freq, na.rm = TRUE),
    Median = median(Freq, na.rm = TRUE),
    Minimum = min(Freq, na.rm = TRUE),
    Maximum = max(Freq, na.rm = TRUE)
  )

# Display the summary table using kable
correlation_summary %>% 
  kbl(caption = "Summary of miRNA Correlations") %>% 
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", 
                                                          "condensed"))

```

The summary of miRNA correlations indicates a total of 261,302 positive correlations and 131,200 negative correlations among the miRNAs. The mean correlation value is 0.40, suggesting a general tendency toward positive correlations, while the median is 0.63, highlighting a skew toward stronger positive relationships. The minimum correlation value is -0.91, showing the presence of strong negative correlations, while the maximum is nearly 1, reflecting near-perfect positive correlations. This distribution underscores the complex and diverse interactions among miRNAs, with both cooperative and antagonistic regulatory relationships evident.


```{r, include=FALSE}
countdata <- data
coldata <- tibble(id=c("Ctrl1", "Ctrl2", "Ctrl3", "Ctrl4", "GBM1", "GBM2", "GBM3", "GBM4", "GBM5", "GBM6", "GBM7", "GBM8", "GBM9"), condition=c("control", "control", "control", "control", "GBM", "GBM", "GBM", "GBM", "GBM", "GBM", "GBM", "GBM", "GBM"))
dds <- DESeqDataSetFromMatrix(countData = countdata,
                              colData = coldata,
                              design= ~ condition, tidy=TRUE)
```

```{r, include=FALSE}
dds <- DESeq(dds)
res <- results(dds)
```

#### 1. Identifying top 10 miRNAs with largest absolute log fold change

```{r}
# Select top 10 miRNAs with largest absolute value of log fold change
res$abs_log2FoldChange <- abs(res$log2FoldChange)
res <- res[order(res$abs_log2FoldChange, decreasing = TRUE),]
data.frame(res[1:10,c(1,2,5)])
```

Difference in miRNA expression level across the control group and GBM group was further investigated using a log2 fold change transformation. Fold change is a measure of quantity change by taking the ratio between the original measurement and subsequent measurement, which in this dataset is the ratio between the GBM group and control group. Thus, a positive value would indicate a large percentage increase in expression in the GBM group compared to the control group, while a negative value would indicate a decrease in expression. Higher absolute values indicate larger changes.

It can be observed that miRNAs that show large changes in expression level generally have a high base expression level. Changes can be as high as 36 times the base level, and p-values are generally small.

```{r}
par(mfrow=c(2,3))

plotCounts(dds, gene="hsa-miR-137|0", intgroup="condition")
plotCounts(dds, gene="hsa-miR-129-2-3p|0", intgroup="condition")
plotCounts(dds, gene="hsa-miR-124-3p|0", intgroup="condition")
plotCounts(dds, gene="hsa-miR-1246|0.005", intgroup="condition")
plotCounts(dds, gene="hsa-miR-218-5p|0", intgroup="condition")
plotCounts(dds, gene="hsa-miR-21-5p|0", intgroup="condition")

```

```{r}
par(mfrow=c(2,3))

plotCounts(dds, gene="hsa-miR-129-5p|0", intgroup="condition")
plotCounts(dds, gene="hsa-miR-128-3p|0.006", intgroup="condition")
plotCounts(dds, gene="hsa-miR-139-3p|0", intgroup="condition")
plotCounts(dds, gene="hsa-miR-433-3p|0", intgroup="condition")


```

For the top ten miRNAs with largest change in expression level, a comparison between control and GBM groups is shown above.

#### 2. Volcano plot

```{r}
# Volcano plot
#reset par
par(mfrow=c(1,1))
# Make a basic volcano plot
with(res, plot(log2FoldChange, -log10(pvalue), pch=20, main="Volcano plot", xlim=c(-3,3)))

# Add colored points: blue if padj<0.01, red if log2FC>1 and padj<0.05)
with(subset(res, padj<.01 ), points(log2FoldChange, -log10(pvalue), pch=20, col="blue"))
with(subset(res, padj<.01 & abs(log2FoldChange)>2), points(log2FoldChange, -log10(pvalue), pch=20, col="red"))
```

Volcano plots are scatterplots that plots significance versus fold change on the y and x axis to quickly identify change in large datasets. In this plot, significance is represented by -log10(p-value) in the y-axis, where points with smaller p-values are expected to have larger values. Change size is indicated by (log2 fold change) on the x-axis, where negative values represent decrease and positive values represent increase. As displayed in the volcano plot, the distribution is slightly asymmetrical, where large decreases in expression level tend to have smaller p-values.

#### 3. PCA

```{r}
vsdata <- vst(dds, blind=FALSE, nsub = nrow(dds))
plotPCA(vsdata, intgroup="condition")
```

As displayed in the PCA plot, the x-axis (PC1) captures the largest source of variability and explains 48% of the total variance in the dataset, while the y-axis (PC2) explains 12% of the variance, representing the second largest, orthogonal source of variability. Together, PC1 and PC2 explain 60% of the total variance. Points representing the control group are tightly clustered on the left side of the plot (negative PC1 values), while points representing the GBM group is clustered on the right. PC1 primarily separates the two groups, suggesting that the variability captured by PC1 is strongly associated with differences between the control and GBM samples. Points are clustered tightly within each group, indicating low variability within groups along both PC1 and PC2.

#### 4. Heat map

```{r}
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds)[,c("condition")])
rownames(df) <- colnames(dds)
pheatmap(assay(dds), cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)
```
